Reasonably Programmable Literal Notation

### 2018-06-22 icfp18/syntax-icfp18.tex
added HTML case study

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    };
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
\label{fig:regex-tlm-def}
\end{subfigure}

@@

TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$re_expr} to support antiquotation across notational boundaries. For example, if \li{$re_expr} had instead used some other antiquotation notation, e.g. \li{^(}$e$\li{)}, then consider this example:
\begin{lstlisting}[numbers=none]
  $proto_expr `( SURLleEURLSURLt y = ^(EURLr1SURL); $m `(...{^(r2)}...)`EURL )`
\end{lstlisting}
The sequence \li{^(r1)} is in expression position, so antiquotation occurs. However, there is no way for \li{$re_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation. In other words, TLM expansion is fundamentally ``outside-in''.  Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender). However, to avoid confusion, we decided not to support the generation of TLM application forms via \li{$re_expr}, and instead repurposed that notation for antiquotation.

This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason. In fact, this is already possible using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below. %(Nested TLM application via splicing is, of course, useful and fully supported, e.g. in Fig.~\ref{fig:nesting-example}).
 % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a

\subsubsection{Correctness Criteria}
\label{sec:correctness-criteria}
In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:
\begin{enumerate}
  \item First, the segmentation is computed and validated. This involves checking two criteria:
    \begin{enumerate}
      \item The segments must be in-bounds, of positive extent, non-overlapping and separated by at least one character.
      \item The segment types must encode valid types assuming only the TLM dependencies.
    \end{enumerate}
  \item Second, the proto-expansion is typechecked assuming only the TLM dependencies (plus opened \li{Pervasives}), treating each splice reference as a variable of the specified segment type.
\end{enumerate}
If validation fails, the client is notified that the applied TLM is incorrectly implemented.% The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.

@@

The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms that contain explicit type annotations. There are various workarounds, e.g. by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step.

Overall, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned. With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin. 
TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve). %MetaOCaml-BER does not change the parse tree or bytecode representation of the OCaml system, and values of type \li{'a code} are not part of the public interface of a TLM parser, so it should, in principle, be possible to compile a parser using MetaOCaml-BER for use in  ``vanilla'' OCaml (and therefore Reason) programs. However, MetaOCaml-BER does not suitably package the necessary pieces of its run-time system, so we have not verified this conjecture.

\subsubsection{Other Implementation Strategies}
\label{sec:html-example}
Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir: 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation. It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.

>>>>

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    };
  };
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
\label{fig:regex-tlm-def}
\end{subfigure}

@@

TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$re_expr} to support antiquotation across notational boundaries. For example, if \li{$re_expr} had instead used some other antiquotation notation, e.g. \li{^(}$e$\li{)}, then consider this example:
\begin{lstlisting}[numbers=none]
  $proto_expr `( SURLleEURLSURLt y = ^(EURLr1SURL); $m `(...{^(r2)}...)`EURL )`
\end{lstlisting}
The sequence \li{^(r1)} is in expression position, so antiquotation occurs. However, there is no way for \li{$re_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation. In other words, TLM expansion is fundamentally ``outside-in''.  Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$re_expr}.

This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason. In fact, this is already available using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below. %(Nested TLM application via splicing is, of course, useful and fully supported, e.g. in Fig.~\ref{fig:nesting-example}).
 % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a

\subsubsection{Correctness Criteria}
\label{sec:correctness-criteria}
In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:
\begin{enumerate}
  \item First, the segmentation is computed and validated. This involves checking two criteria:
    \begin{enumerate}
      \item The segments must be in-bounds, of positive extent, non-overlapping and separated.
      \item The segment types must encode valid types assuming only the TLM dependencies.
    \end{enumerate}
  \item Second, the proto-expansion is typechecked assuming only the TLM dependencies (plus opened \li{Pervasives}), treating each splice reference as a variable of the specified segment type.
\end{enumerate}
If validation fails, the client is notified that the applied TLM is incorrectly implemented.% The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.

@@

The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms with type annotations. One workaround is to define for each  segment type \li{t} a dummy value of type \li{t code} that is replaced with a splice reference by post-processing after the coercion step.

Overall, this approach can, at least, substantially increase a TLM provider's confidence in the type correctness of the expansion logic. 
% With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin. 
TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve). %MetaOCaml-BER does not change the parse tree or bytecode representation of the OCaml system, and values of type \li{'a code} are not part of the public interface of a TLM parser, so it should, in principle, be possible to compile a parser using MetaOCaml-BER for use in  ``vanilla'' OCaml (and therefore Reason) programs. However, MetaOCaml-BER does not suitably package the necessary pieces of its run-time system, so we have not verified this conjecture.

\subsubsection{Other Implementation Strategies}
\label{sec:html-example}

\begin{figure}[t]
\begin{subfigure}[t]{0.46\textwidth}
\begin{lstlisting}
  module Html = {
    /* a simplified encoding of HTML */
    type tag = string;
    type attr = (string, string);
    type attrs = list(attr);
    type t = 
      | Text(string) 
      | Elem(tag, attrs, list(t));
  };
\end{lstlisting}
\vspace{-6px}
\caption{The \li{Html} module, which defines \li{Html.t}.}
\label{fig:html-type-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.52\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module HtmlNotation = {
    notation $html at Html.t {
      lexer HtmlLexer
      expression parser HtmlParser.expr
      pattern parser HtmlParser.pat 
      in package html_parser;
      dependencies = {module Html = Html}
    };
  };
\end{lstlisting}
\vspace{-6px}
\caption{The \li{$html} expression and pattern TLMs.}
\label{fig:html-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{3px}
% let pElem : Html.tag -> Html.attrs -> list(ProtoExpr.t) -> ProtoExpr.t = (* ... *)
% let pTextElem : Html.tag -> Html.attrs, string) -> ProtoExpr.t = (* ... *)
\begin{lstlisting}
  module HtmlLexer = Relit.TrivLexer;
  module HtmlParser = Relit.HandRolledExprPatParser({
    module ProtoHtml = {
      type t = PText(string)
             | PElem(Html.tag, Html.attrs, list(t))
             | PSplicedElem(Relit.Segment.t) 
             | PSplicedText(Relit.Segment.t) 
      let parse : string => t = 
        /* ... via Markup.ml's stream combinators (see text) ... */;
    };
    module ExprExpander = {
      open notation Relit.$proto_expr; open ProtoHtml;
      let expand_attr = (x1, x2) => `(SURL ($s `(EURLx1SURL)`, $s `(EURLx2SURL)`)EURL )`;
      let expand_attrs = (attrs) => `(SURL $list `(EURLList.map(expand_attr, attrs)SURL)`EURL )`;
      let expand_proto = fun 
      | PText(text) => `(SURL Html.Text($s `(EURLtextSURL)`)EURL )` 
      | PElem(tag, attrs, children) => 
        `(SURL Html.Elem($s `(EURLtagSURL)`, `(EURLexpand_attrs(attrs)SURL)`, 
            $list `(EURLList.map(expand_parsed, children)SURL)`)EURL )`
      | PSplicedHtml(seg) => `(SURL $spliced `(EURLsegSURL : Html.t) EURL)`);
      | PSplicedText(seg) => `(SURL Html.Text($spliced `(EURLsegSURL : striSURLEURLng)`) EURL)`
    };
    let expr = (body) => body |> ProtoHtml.parse |> ExprExpander.expand_proto;
    let pat  = (body) => expr(body) |> ProtoPat.from_expr;
  })
\end{lstlisting}
\vspace{-6px}
\caption{The implementation of \li{HtmlParser}, which defers the parsing step to the \li{Markup.ml} library \cite{markupml}.}
\label{fig:html-parser}
\vspace{-5px}
\end{subfigure}
% \begin{subfigure}[t]{\textwidth}
% \lstset{deletestring=[d]{"}}
% \begin{lstlisting}
%   open HtmlNotation; 
%   exception UnexpectedFormat;
%   let scrape_data = $html.(fun 
%   | `(SURL<div... class="result">EURL
%       [ `(SCOLOR<h1><$$>ECOLORtitleSCOLOR</$$></h1>ECOLOR)`, author, summary, ..._ ]SURL</div>EURL)` => 
%     (title, author, summary)
%   | _ => raise UnexpectedFormat);
% \end{lstlisting}
% \vspace{-4px}
% \caption{Using a pattern TLM to deconstruct an HTML value.}
% \label{fig:pattern-tlm-example}
% \end{subfigure}
\vspace{5px}
\caption{Case Study: HTML literals as a library}
\vspace{-4px}
\label{fig:html-case-study}
\end{figure}

Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir. 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
% the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. 
%For example, the accompanying technical report briefly describes how \li{$re_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation. 
Let us consider one substantial example of this approach here: the \li{$html} TLM, shown being applied in Fig.~\ref{fig:nesting-example}, is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step as outlined in Fig.~\ref{fig:html-case-study}.

There are many ways to represent HTML data, but one simple representation is specified by \li{Html.t}, defined in Fig.~\ref{fig:html-type-def}. However, manually applying the constructors is tedious and, for many web programmers, the induced notation is unfamiliar. It also makes it difficult to copy-and-paste from, for example, existing HTML files, as might be useful when refactoring an existing project to use Reason. 
The \li{$html} TLM defined in Fig.~\ref{fig:html-tlm-def} implements the standard HTML notation extended with the tags \li{<$>} and \li{<$$>}, which support \li{Html.t} splicing and \li{string} splicing, respectively. 
%It also supports the suffix \li{...} on a tag, which causes the body of the tag to be a splice of type \li{list(Html.t)}. 
{We would also in practice have splicing for HTML lists and attribute values, but we omit these here.}

Fig.~\ref{fig:html-parser} shows much of the implementation of \li{HtmlParser},  which is constructed by applying a functor that constructs a module that has the same essential interface as the modules that Menhir generates, with the functions \li{expr : string -> ProtoExpr.t} and \li{pat : string -> ProtoPat.t} of the input module exported as the ``non-terminals''. The companion lexer is \li{Relit.TrivLexer}, which simply passes the whole literal body on as a single token. The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}. (We discuss \li{pat} in the next section\todo{do this}{}.)

The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression. We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, described earlier, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream. Finally, it folds over this stream to produce the result of type \li{ProtoHtml.t}.

The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew. Overall, the amount of code needed to implement HTML literals.

Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React). The expansion of this notation must be interpretted by a PPX rewriter. The problem is that this makes it difficult to use different HTML-related libraries at once. For example, ReasonReact and an XML library might both want to use this notation. This approach solves this problem (as well as other problems with the PPX-based approach as discussed in Sec.~\ref{sec:existing-approaches}).
